# Lab 2: Basic R Programming and Vector Operations {#lab2}

```{block, type='objectives'}

**Objectives:**

1. To understand what an object is in R
2. To identify and create vectors
3. To learn how to differentiate types (classes) of vectors
4. To perform simple math operations on numeric vectors
```

This chapter focuses on getting comfortable with the `R` console, scripting, and packages, and on the core syntax you will use for the rest of the course: creating objects, and building and operating on vectors.

## The R console and basic math

Open RStudio and locate the **console** pane. You can type any of the following directly into the console and hit `Enter`:

```{r}
2 + 2
10 - 3
6 * 7
100 / 4
2^5
```

R follows the standard order of operations (PEMDAS), and you can use parentheses to control it, just like a calculator.

```{block, type='rmdquestion'}
#### Question 1
   - Use the R console to calculate the area of a circle with radius 4.5 (Hint: `pi` is a built-in constant in R)
   - Calculate `(12+8)/4^2` using R. Add the code and the result.
```

## Objects

`R` can store any type of variable into an **object**. To create an object, write the name of the object, followed by `<-`, followed by what you want to store.

```{r}
x <- 5
x
```

You'll see the new object `x` appear in your **Environment** pane (upper right of RStudio). That confirms your object has been created and is available for future use in this session.

## Basic data structures in R

A basic data structure is an n-dimensional object that stores information. Data structures in R are classified by their number of dimensions:

1. Uni-dimensional data structures such as **vectors**
2. Two-dimensional data structures such as **data frames** or **matrices**
3. N-dimensional data structures such as **lists** (not covered in this course)

```{r}
# One dimensional objects
c(1,2,3,4,5)
c("A","B","C","D","E")

# Two dimensional objects
data.frame(c(1,2,3,4,5), c("A","B","C","D","E"))
```

## Unidimensional objects: Vectors

A vector is a one-dimensional object — think of it as a "list" of elements in only one direction, with no rows or columns.

```{r}
numbers <- c(1,2,3,4,5)
letters_vec <- c("A","B","C","D")
numbers
letters_vec
```

To create a vector, use the `c()` function (short for "combine"), which combines the values you separate by commas.

```{block, type='rmdinfo'}
###  R Functions

Functions are the commands `R` uses to interpret an order. `c()` means "Combine Values into a Vector."

Syntax: `function(object)`

To measure the length of a vector, use the function `length()`:
```

```{r}
length(numbers)
```

```{block, type='rmdquestion'}
#### Question 2

- Create three vectors:
    - One named `words` with your five favorite words
    - One named `numbers` with your lucky numbers
    - One named `mixed` with three numbers and three letters, in any order
- Show the code you used to create the vectors
```

### Classes of vectors

Use the `class()` command to identify the type of vector you're working with:

```{r}
class(numbers)
class(letters_vec)
```

`R` shows either `numeric`/`integer` or `character`. Character vectors are surrounded by quotation marks (`"`); numeric vectors are not.

```{block, type='rmdquestion'}
#### Question 3

- What are the classes of your `words`, `numbers` and `mixed` objects from Question 2?
- What class does R assign to a vector that mixes numbers and letters, like `mixed`? Why do you think that happens?
```

## Math operations on vectors

Because vectors can hold multiple numbers at once, math operations are applied **element-wise**:

```{r}
v1 <- c(1, 2, 3, 4, 5)
v1 + 10
v1 * 2
v1 ^ 2
```

You can also do math between two vectors of the same length:

```{r}
v2 <- c(10, 20, 30, 40, 50)
v1 + v2
v2 / v1
```

Useful summary functions for numeric vectors:

```{r}
sum(v1)
mean(v1)
max(v1)
min(v1)
```

```{block, type='rmdquestion'}
#### Question 4

- Create a numeric vector called `measurements` with 6 numbers of your choice
- Calculate the sum, mean, maximum, and minimum of `measurements`. Add the code and results
- Multiply every value in `measurements` by 2.54 (as if converting inches to centimeters). Add the code and the resulting vector
```

## Expanding vectors

If you have two vectors of the same class, you can combine them using `c()` again:

```{r}
double.letters <- c(letters_vec, letters_vec)
double.letters

double.numbers <- c(numbers, numbers)
double.numbers

more.numbers <- c(numbers, 6, 7, 8, 9, 10)
more.numbers
```

```{block, type='rmdquestion'}
#### Question 5

- Increase each of your vectors from Question 2 by two or three more elements. Add the code and the results.
- Create an object that combines your `words` and `numbers` vectors. Add the code.
- What is the class of this combined object? Explain why, based on what you learned in Question 3.
```
