# Lab 4: Scientific Visualization with ggplot2 {#lab4}

```{block, type='objectives'}

**Objectives:**

1. To understand the Grammar of Graphics used by ggplot2
2. To map variables to aesthetics (x, y, fill, color)
3. To build histograms and bar plots
4. To design a publication-quality chart
```

One of the clearest ways to present results is by plotting them. This chapter focuses on the `ggplot2` package, the most widely used plotting package in `R`.

## The Grammar of Graphics

`ggplot2` builds plots in layers using a consistent syntax:

```
ggplot(data = <data>, aes(x = <variable>, y = <variable>)) + geom_<type>()
```

- `data =` tells ggplot which data frame to use
- `aes()` ("aesthetic mapping") tells ggplot which columns go on which axis (and optionally, color/fill)
- `geom_<type>()` tells ggplot what kind of plot to draw (bars, points, lines, etc.)

Make sure your data is "tidy": each variable you want to plot should be in its own column.

## Installing and loading ggplot2

```{block, type='rmdinfo'}
#### R Packages

An R package is a collection of code, data, and functions that extends what `R` can do.

##### Installing a package

Write this once in the **CONSOLE**:

`install.packages('ggplot2')`

##### Loading a package

Add this to a code chunk in your R Markdown notebook:

\```{r}
library(ggplot2)
\```
```

We'll practice with `R`'s well-known `palmerpenguins` data set. Install it if you haven't already (`install.packages("palmerpenguins")`), then load both packages:

```{r}
library(ggplot2)
library(palmerpenguins)
data(penguins)
```

```{block, type='rmdquestion'}
#### Question 1

- Load `ggplot2` and `palmerpenguins` in your own notebook
- Use `str(penguins)` or `head(penguins)` to explore the data. What are the columns, and what class is each one?
```

## Histograms

A histogram shows the **distribution of frequency** of a continuous value. The x-axis holds the continuous values (binned into intervals) and the y-axis holds the count.

```{r}
ggplot(data = penguins, aes(x=flipper_length_mm)) + geom_bar()
```

We can add color to separate categories, for example by species:

```{r}
ggplot(data = penguins, aes(x=flipper_length_mm, fill=species)) + geom_bar()
```

```{block, type='rmdquestion'}
#### Question 2

- Make a histogram of `bill_length_mm`, colored by `species`. Add the code and a short interpretation: do the species overlap or separate?
```

## Bar plots

Bar plots represent a **continuous variable across discrete categories**. A good example: the number of penguins per species per island.

```{r}
library(tidyverse)
penguin.island <- penguins %>% count(species, island)
penguin.island
```

```{r}
ggplot(data = penguin.island, aes(x=island, y=n)) + geom_bar(stat = "identity")
```

The `stat="identity"` flag tells ggplot to use the actual value in `n` for the bar height, instead of counting rows.

Add `fill=species` to break each bar down by species:

```{r}
ggplot(data = penguin.island, aes(x=island, y=n, fill=species)) + geom_bar(stat = "identity")
```

This is a **stacked bar chart**. To place bars for each category side-by-side instead, add `position="dodge"`:

```{r}
ggplot(data = penguin.island, aes(x=island, y=n, fill=species)) + geom_bar(stat = "identity", position = "dodge")
```

```{block, type='rmdquestion'}
#### Question 3

- Which islands have only one species of penguin, and which have more than one? Use the plots above to answer.
- Recreate the `penguin.island` summary but grouped by `species` and `sex` instead of `species` and `island`. Plot it as a dodged bar chart.
```

## Publication-quality touches

A few small additions go a long way toward making a plot presentation-ready:

```{r}
ggplot(data = penguin.island, aes(x=island, y=n, fill=species)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Penguin counts by island and species",
       x = "Island", y = "Number of penguins", fill = "Species") +
  theme_minimal()
```

- `labs()` sets a title and axis/legend labels
- `theme_minimal()` (or `theme_bw()`, `theme_classic()`) strips ggplot's default gray background

```{block, type='rmdquestion'}
#### Question 4

- Take one of your plots from this lab and add a title, clear axis labels, and a theme of your choice
- In 2-3 sentences, explain why labeling and theming matters when sharing a plot outside of your own notebook (e.g., in a paper or presentation)
```
