# Lab 5: Chaining Data Operations and Pipelines for Public Health Data {#lab5}

```{block, type='objectives'}

**Objectives:**

1. To understand the purpose of the pipe operator (`%>%` or `|>`)
2. To chain multiple functions into a single readable pipeline
3. To structure clean, reproducible code for multi-step analyses
4. To apply a pipeline to a public health data set
```

So far, every time we wanted to do multiple operations on our data, we either nested functions inside each other or created many intermediate objects. Today we learn a cleaner way: **piping**.

## The problem with nested functions

Recall from earlier labs that we can nest functions, like `length(unique(x))`. This works for two steps, but gets hard to read fast:

```{r, eval=FALSE}
round(mean(subset(my_data, group == "A")$value), 2)
```

Reading this requires working from the inside out — the opposite order of how the steps actually happen.

## The pipe operator

The pipe operator, written `%>%` (from the `tidyverse` package) or `|>` (built into base R since version 4.1), means **"and then."** It takes the output of what's on its left and feeds it as the first input to what's on its right.

```{r}
library(tidyverse)
```

```{r}
c(4, 8, 15, 16, 23, 42) %>% mean()
```

is the same as:

```{r}
mean(c(4, 8, 15, 16, 23, 42))
```

The real benefit shows up once you chain several steps:

```{r}
c(4, 8, 15, 16, 23, 42) %>%
  mean() %>%
  round(1)
```

Each line reads like a sentence: "take this vector, and then take its mean, and then round it."

```{block, type='rmdquestion'}
#### Question 1

- Rewrite `sqrt(sum(c(3,4)^2))` (the length of a right-triangle hypotenuse) using the pipe operator
- In your own words, explain what "and then" means in the context of `%>%`
```

## Building a pipeline on a data frame

Let's build a small public health data set: daily case counts of a disease across a few counties.

```{r}
cases <- data.frame(
  county = c("Worcester","Worcester","Middlesex","Middlesex","Suffolk","Suffolk"),
  day = c(1,2,1,2,1,2),
  new_cases = c(12, 18, 5, 9, 30, 25),
  stringsAsFactors = FALSE
)
cases
```

A typical pipeline might: filter to a county of interest, calculate a summary statistic, and round the result — all in one chain:

```{r}
cases %>%
  filter(county == "Suffolk") %>%
  summarize(total_cases = sum(new_cases), avg_cases = mean(new_cases))
```

`filter()` and `summarize()` are `tidyverse` functions similar to `subset()` and the summary functions you've already used, but designed to work smoothly with the pipe.

We can also group by a category before summarizing, using `group_by()`:

```{r}
cases %>%
  group_by(county) %>%
  summarize(total_cases = sum(new_cases))
```

```{block, type='rmdquestion'}
#### Question 2

- Add two more days of data to the `cases` data frame (for any counties)
- Write a pipeline that calculates the average `new_cases` per county, sorted from highest to lowest (hint: look up the `arrange()` function and use `desc()`)
- Add the code and the resulting table
```

## Adding a new calculated column mid-pipeline

The `mutate()` function adds a new column, and can be chained into your pipeline:

```{r}
cases %>%
  mutate(cases_per_1000 = new_cases / 10) %>%
  filter(cases_per_1000 > 1)
```

```{block, type='rmdquestion'}
#### Question 3

- Add a column to `cases` called `severity`, calculated as `new_cases * 2` (a made-up severity index), using `mutate()` inside a pipeline
- Then filter the pipeline to show only rows with `severity` greater than 20
- Chain all of this into a single pipeline with `%>%`. Add the code and result
```

## Why pipelines matter for reproducibility

A well-structured pipeline documents your entire analysis as a readable sequence of steps, each easy to check individually. This connects directly to the goal from Lab 1: an outside reader (or your future self) should be able to follow exactly what happened to the data, in what order.

```{block, type='rmdquestion'}
#### Question 4

- Take any 3-step analysis you've done in a previous lab (loading data, filtering, and summarizing, for example) and rewrite it as a single `%>%` pipeline
- What are the advantages of a pipeline over nested functions, in terms of reproducibility and readability? What might be a disadvantage?
```
