# Lab 8: Advanced Data Wrangling and Importing Ecological Datasets {#lab8}

```{block, type='objectives'}

**Objectives:**

1. To read external data files (.txt, .csv, .xlsx) into R
2. To identify and handle missing data
3. To select specific variables and add new calculated columns
4. To summarize data by category
```

Real ecological data rarely arrives ready to analyze. This lab covers how to bring outside files into `R`, clean them up, and reshape them into something usable.

## Reading text-delimited files (`.txt` / `.tsv`)

**Text delimited files** contain plain, unformatted text, where columns are separated by a `tab` character (`\t`). To read them, use `read.table()`:

```{r}
# read.table(file = "data_table.txt", sep = "\t", header = T)
```

```{block, type='rmdquestion'}
#### Question 1

- Download the table from [here](https://raw.githubusercontent.com/Tabima/QuantBio_BIO120/main/Lab_5/data_table.txt) and read it into `R`. Add the code.
- What do the `sep` and `header` options mean? (Use `?read.table` to check)
- What happens if you remove the `sep` and `header` options?
```

## Reading comma-separated files (`.csv`)

CSV files are the same idea, but columns are separated by commas instead of tabs. Use `read.csv()`:

```{block, type='rmdquestion'}
#### Question 2

- Download the table from [here](https://raw.githubusercontent.com/Tabima/QuantBio_BIO120/main/Lab_5/data_table.csv) and read it into `R` using `read.csv()`. Add the code and an explanation of the syntax.
- Is this table any different from the one in Question 1?
```

## Reading Excel files (`.xlsx`)

Excel files are extremely common in the biological sciences, especially from collaborators who aren't programmers. Use the `read_xlsx()` function from the `readxl` package:

```{block, type='rmdquestion'}
#### Question 3

- Install the `readxl` package
- Download the excel spreadsheet from [here](https://raw.githubusercontent.com/Tabima/QuantBio_BIO120/main/Lab_5/data_table.xlsx)
- Using `read_xlsx()`, read the sheets into two separate objects (check your spreadsheet's tab names first)
- Is this table any different from the ones in Questions 1 and 2?
```

## Dealing with missing data

Ecological field data almost always has gaps — a sensor failed, a sample was lost, an observer skipped a field. `R` represents missing values as `NA`.

```{r}
eco_data <- data.frame(
  site = c("A","B","C","D","E"),
  temp_c = c(18.2, NA, 21.4, 19.8, NA),
  ph = c(7.1, 6.9, NA, 7.3, 7.0)
)
eco_data
```

Check for missing values with `is.na()`:

```{r}
is.na(eco_data)
```

Count how many missing values are in a column:

```{r}
sum(is.na(eco_data$temp_c))
```

You have a few options for handling missingness:

```{r}
# Remove any row with at least one NA
na.omit(eco_data)

# Calculate a mean while ignoring NAs
mean(eco_data$temp_c, na.rm = TRUE)
```

```{block, type='rmdquestion'}
#### Question 4

- Using the `eco_data` example (or a similar data frame of your own with at least 8 rows and some `NA` values), calculate how many missing values exist in each numeric column
- Calculate the mean of each numeric column while ignoring `NA`s
- In your own words: what is the risk of simply deleting every row with a missing value (`na.omit`), versus keeping the row and ignoring the `NA` in your calculations (`na.rm=TRUE`)?
```

## Selecting variables and adding calculated columns

Recall `subset()` from Lab 3 for filtering rows. The `select` argument lets you keep or drop specific **columns**:

```{r}
# Keep only site and temp_c
subset(eco_data, select = c(site, temp_c))

# Drop the ph column
subset(eco_data, select = -c(ph))
```

To add a new calculated column, you can assign directly with `$`, or use `mutate()` from `tidyverse` (Lab 5):

```{r}
eco_data$temp_f <- (eco_data$temp_c * 9/5) + 32
eco_data
```

```{block, type='rmdquestion'}
#### Question 5

- Remove the `ph` column from your own ecological data set and explain why you might want to do this in a real analysis
- Add a new calculated column of your choice (a unit conversion, an index, a rounded value, etc.) using either `$` or `mutate()`. Add the code and the resulting table.
```

## Summarizing data by category

Combine what you learned in Lab 5 (`group_by`, `summarize`) with today's cleaned-up data to produce site-level summaries:

```{r}
library(tidyverse)
eco_data %>%
  group_by(site) %>%
  summarize(mean_temp = mean(temp_c, na.rm = TRUE))
```

```{block, type='rmdquestion'}
#### Question 6

- Using one of the data sets you imported in Questions 1-3, group by a categorical column and summarize a numeric column (mean, min, max, or count). Add the code and result.
- What ecological or biological question could this summary help answer?
```
