# Lab 9: Advanced ggplot2 - Facets, Themes, and Layouts {#lab9}

```{block, type='objectives'}

**Objectives:**

1. To build scatterplots for continuous-vs-continuous relationships
2. To build boxplots and violin plots for distributions across categories
3. To refine axes, scales, and custom themes
4. To use facetting to display complex, multi-panel data sets
```

In Lab 4 we covered histograms and bar plots. Today we go further: relationships between continuous variables, distributions across groups, and multi-panel layouts — the kind of polish you'll want for a scientific paper or presentation.

## Scatterplots

Scatterplots represent **continuous vs. continuous** variables, and are used to look for correlations.

```{r}
library(ggplot2)
library(palmerpenguins)
data(penguins)

ggplot(data = penguins, aes(x=bill_depth_mm, y=bill_length_mm)) + geom_point()
```

This is hard to read as-is — no clear trend appears. Coloring by a category often reveals a hidden pattern (this is called **Simpson's Paradox**, when a trend appears, disappears, or reverses when a grouping variable is introduced):

```{r}
ggplot(data = penguins, aes(x=bill_depth_mm, y=bill_length_mm, color=species)) + geom_point()
```

```{block, type='rmdquestion'}
#### Question 1

- Build a scatterplot of `flipper_length_mm` (x) versus `body_mass_g` (y), colored by `species`
- Is there a positive, negative, or no relationship? Does it differ by species?
```

## Boxplots and violin plots

These combine a histogram-like view with a categorical x-axis, illustrating the **distribution of a continuous variable across categories**.

```{r}
ggplot(data = penguins, aes(x=species, y=flipper_length_mm, color=species)) + geom_violin()
```

Add the raw data points with `geom_jitter()`:

```{r}
ggplot(data = penguins, aes(x=species, y=flipper_length_mm, color=species)) + geom_violin() + geom_jitter()
```

Add a boxplot on top to summarize the median and quartiles:

```{r}
ggplot(data = penguins, aes(x=species, y=flipper_length_mm, color=species)) + geom_violin() + geom_jitter() + geom_boxplot()
```

The thick middle line is the median; the box edges are the 25th and 75th percentiles; the whiskers approximate a 95% range for comparing medians.

```{block, type='rmdquestion'}
#### Question 2

- Build a boxplot of `bill_length_mm` across `species`
- Which species has the widest spread (most variable) bill length? Which has the narrowest?
- Connect this back to Lab 6: if you ran an ANOVA on this variable, would you predict a significant result based on this plot alone? Why?
```

## Facetting: multiple panels for complex data

Facetting splits one plot into a grid of small panels, one per category — extremely useful for comparing patterns across many groups at once (e.g., decadal climate trends across multiple regions).

```{r}
ggplot(data = penguins, aes(x=bill_depth_mm, y=bill_length_mm)) +
  geom_point() +
  facet_wrap(~species)
```

`facet_wrap()` splits by one variable. `facet_grid()` splits by two variables at once (rows and columns):

```{r}
ggplot(data = penguins, aes(x=bill_depth_mm, y=bill_length_mm)) +
  geom_point() +
  facet_grid(sex ~ species)
```

```{block, type='rmdquestion'}
#### Question 3

- Facet the boxplot from Question 2 by `island` using `facet_wrap()`
- Does the pattern you observed in Question 2 hold across all islands, or does it change? Explain.
```

## Refining axes, scales, and themes

```{r}
ggplot(data = penguins, aes(x=species, y=flipper_length_mm, fill=species)) +
  geom_boxplot() +
  scale_y_continuous(breaks = seq(170, 235, by = 10)) +
  labs(title = "Flipper length by species",
       subtitle = "Palmer Station, Antarctica LTER",
       x = "Species", y = "Flipper length (mm)") +
  theme_bw() +
  theme(legend.position = "none")
```

- `scale_y_continuous(breaks = ...)` controls exactly where axis tick marks appear
- `theme_bw()` gives a clean white-background theme, common in publications
- `theme(legend.position = "none")` removes a redundant legend (here, color duplicates the x-axis)

```{block, type='rmdquestion'}
#### Question 4

- Take your faceted plot from Question 3 and add a title, subtitle, clean axis labels, and a theme of your choice
- Compare it to the very first, unstyled scatterplot at the top of this lab. In 2-3 sentences, explain what specifically makes the final version more "publication-ready."
```
