# Lab 11: Computational Data Fitting and Predictive Modeling {#lab11}

```{block, type='objectives'}

**Objectives:**

1. To fit a linear regression model in R
2. To interpret model coefficients, R-squared, and p-values
3. To evaluate model fit using residuals
4. To use a fitted model to forecast/predict new values
```

This is our final statistics lab of the semester. Instead of just testing whether groups differ (Labs 6 and 10), today we build a model that describes **how** one variable changes as another changes — and use it to make predictions, a common task in environmental and conservation biology (e.g., forecasting population trends or growth curves).

## Linear regression basics

A simple linear regression models a straight-line relationship:

$$y = \beta_0 + \beta_1 x + \epsilon$$

where $\beta_0$ is the intercept, $\beta_1$ is the slope, and $\epsilon$ is the error term.

Let's use body measurements from the `penguins` data set (Lab 4/9) to model how flipper length predicts body mass.

```{r}
library(ggplot2)
library(palmerpenguins)
data(penguins)
penguins <- na.omit(penguins)
```

```{r}
model <- lm(body_mass_g ~ flipper_length_mm, data = penguins)
summary(model)
```

Key parts of the output:

- **Estimate** (Intercept and slope): the fitted values of $\beta_0$ and $\beta_1$
- **Pr(>|t|)**: p-value testing whether each coefficient differs significantly from zero
- **R-squared**: the proportion of variance in $y$ explained by $x$ (0 to 1; closer to 1 = better fit)

```{block, type='rmdquestion'}
#### Question 1

- What is the slope of this model? In plain language, what does it mean biologically (e.g., "for every 1mm increase in flipper length, body mass increases by ___ grams")?
- What is the R-squared value? How much of the variation in body mass does flipper length explain?
- Is the slope statistically significant? How do you know?
```

## Visualizing the fitted line

```{r}
ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE) +
  labs(x = "Flipper length (mm)", y = "Body mass (g)") +
  theme_minimal()
```

`geom_smooth(method="lm")` overlays the regression line, and `se=TRUE` shades a 95% confidence band around it.

```{block, type='rmdquestion'}
#### Question 2

- Recreate this plot but colored by `species`. Does the flipper length - body mass relationship look consistent across species, or different?
```

## Checking the fit: residuals

**Residuals** are the differences between observed and predicted values ($y_{observed} - y_{predicted}$). A good linear fit should have residuals scattered randomly around zero, with no obvious pattern.

```{r}
plot(model, which = 1)
```

If you see a clear curve or funnel shape in this plot, a straight line may not be the right model for your data.

```{block, type='rmdquestion'}
#### Question 3

- Based on the residual plot, does a linear model seem like a reasonable fit for this relationship? Explain what you'd look for to answer "yes" or "no."
```

## Making predictions

Once a model is fit, you can use it to predict new, unobserved values with `predict()`:

```{r}
new_penguin <- data.frame(flipper_length_mm = 210)
predict(model, newdata = new_penguin)
```

```{block, type='rmdquestion'}
#### Question 4

- Using your fitted model, predict the body mass of a penguin with a flipper length of 195mm and one with 225mm
- One of these predictions is likely more reliable than the other. Which one, and why? (Hint: look at the range of flipper lengths actually observed in the data with `range(penguins$flipper_length_mm)` — predicting outside this range is called "extrapolation")
```

## Curve fitting for growth and conservation trends

Not all biological relationships are straight lines. Population growth, for example, is often better modeled with a curve. Here's a simple example using `nls()` (nonlinear least squares) to fit an exponential growth curve to simulated population data:

```{r}
set.seed(11)
year <- 1:15
population <- 50 * exp(0.15 * year) + rnorm(15, 0, 10)
pop_data <- data.frame(year, population)

growth_model <- nls(population ~ N0 * exp(r * year), data = pop_data,
                     start = list(N0 = 50, r = 0.1))
summary(growth_model)
```

```{r}
ggplot(pop_data, aes(x = year, y = population)) +
  geom_point() +
  geom_line(aes(y = predict(growth_model)), color = "blue") +
  labs(title = "Fitted exponential growth curve", x = "Year", y = "Population size") +
  theme_minimal()
```

```{block, type='rmdquestion'}
#### Question 5

- What is the fitted growth rate ($r$) from `growth_model`? 
- Use `predict()` with this model to forecast the population size at year 20. Add the code and result.
- Why might a linear model be a poor choice for forecasting population growth, compared to this exponential model? Relate your answer back to what you observed with residuals in Question 3.
```
